---
title: "Project 3 : Markov switching models"
author: 'Mattia Guerri - Hugo Brehier'
subtitle: 'Statistics for Smart Data'
date: 'November 29^th^ 2019'
output: pdf_document
---

\pagebreak
\tableofcontents
\pagebreak


```{r global_options, include=FALSE, echo = FALSE, warning = FALSE, fig.align="center"}
knitr::opts_chunk$set(echo=TRUE, warning=FALSE, message=FALSE)
```

# Introduction

In this project, we study Markov Switching Models. 
First, a simulation study is done.
Next, we define the model and the estimation algorithm.
Finally, we apply it to real financial data.

\pagebreak

# Simulated Data

Install the depmixS4 and quantmod libraries
```{r}
library('depmixS4')
library('quantmod')
library(ggplot2)
library(scales)
set.seed(1)
```

Create the parameters for the bull and bear market returns distributions
```{r}
Nklower <- 50
Nkupper <- 150
bullmean <- 0.1
bullvar <- 0.1
bearmean <- -0.05
bearvar <- 0.2
```

Create the list of durations (in days) for each regime
```{r}
days = replicate(5, sample(Nklower :Nkupper, 1))
```

Create the various bull and bear markets returns
```{r}
marketbull1 <- rnorm( days[1], bullmean, bullvar )
marketbear2 <- rnorm( days[2], bearmean, bearvar )
marketbull3 <- rnorm( days[3], bullmean, bullvar )
marketbear4 <- rnorm( days[4], bearmean, bearvar )
marketbull5 <- rnorm( days[5], bullmean, bullvar )
```

Create the list of true regime states and full returns list
```{r}
trueregimes <-c( rep(1,days[1]), rep(2,days[2]), rep(1,days[3]), rep(2,days[4]), rep(1,days[5]))
returns <-c( marketbull1, marketbear2, marketbull3, marketbear4, marketbull5)
```

Plotting the returns shows the clear changes in mean and variance between the regime
switches.
```{r}
plot(returns, type="l", xlab='', ylab="Returns")
```

 Create and fit the Hidden Markov Model
```{r}
hmm = depmix(returns ~ 1, family = gaussian(), nstates = 2, data=data.frame(returns))
hmmfit = fit(hmm, verbose = FALSE)
```

 Output both the true regimes and the posterior probabilities of the regimes
```{r}
postprobs <- posterior(hmmfit)
layout(1 :2)
plot(postprobs$state, type='s', main='True Regimes', xlab='', ylab='Regime')
matplot(postprobs[, -1], type='l', main='Regime Posterior Probabilities',ylab='Probability')
legend(x='topright', c('Bull','Bear'), fill=1:2, bty='n')
```

\pagebreak
# Model behind the code
## Markov Switching Models

Write the MS model.
Documented from \cite{JH_RMS} and \cite{GT_RMS} and \cite{FA_RMS} . 


## EM algorithm

EM Estimation method, maximum likelihood. Also Backward-Forward ?

\pagebreak
# Application to real financial data
## S&P Index

Obtain S&P500 data from 2004 onwards
```{r}
getSymbols( '^GSPC',src="yahoo", from="2004-01-01")
    ```

```{r}
values = as.data.frame(GSPC)
n= nrow(values)
values$date = as.Date(rownames(values))

ggplot(values, aes(x=date, y=GSPC.Open,group = 1)) + geom_line() + scale_x_date(breaks = date_breaks("years"),
  labels = date_format('%y')) + xlab('year')

```

```{r}
gspc.returns = as.data.frame((values[2:n, 1] - values[1:(n-1), 1]) /  values[1:(n-1), 1])
gspc.returns$date = as.Date(rownames(values)[2:n])

colnames(gspc.returns) = c('returns','date')

ggplot(gspc.returns, aes(x=date, y=returns,group = 1)) + geom_line() + scale_x_date(breaks = date_breaks("years"),
  labels = date_format('%y')) + xlab('Year')
```

```{r}
k=10
AICs = numeric(length(2:k))

for (i in 2:k) {
  hmm <- depmix(returns ~ 1, family = gaussian(), nstates = i, data=data.frame(gspc.returns))
  hmmfit <- fit(hmm, verbose = FALSE)
  AICs[i-1] = AIC(hmmfit)
}

```

```{r}
bestk = which.min(AICs) + 1
bestk
```

```{r}
hmm <- depmix(returns ~ 1, family = gaussian(), nstates = bestk, data=data.frame(gspc.returns))
hmmfit <- fit(hmm, verbose = FALSE)
```

```{r}
summary(hmmfit)
```



```{r}
dat = posterior(hmmfit)
dat$date = gspc.returns$date
dat$ret = gspc.returns$returns
```

```{r}
ggplot(dat, aes(x=date, y=ret,color=factor(state),group = 1)) + geom_line() + scale_x_date(breaks = date_breaks("years"),
  labels = date_format('%y')) + xlab('Year') + scale_color_brewer(palette="Set1")
```

The most volatil bear state appears almost only during the subprime crisis. It shows the market was in an exceptional state.

## IWM

From etf.com:
IWM tracks a market-cap-weighted index of US small-cap stocks. The index selects stocks ranked 1,001-3,000 by market cap.

```{r}
getSymbols( 'IWM',src="yahoo", from="2004-01-01" )
```

```{r}
values = as.data.frame(IWM)
n= nrow(values)
values$date = as.Date(rownames(values))

ggplot(values, aes(x=date, y=IWM.Open,group = 1)) + geom_line() + scale_x_date(breaks = date_breaks("years"),
  labels = date_format('%y')) + xlab('year')
```

```{r}
iwm.returns = as.data.frame((values[2:n, 1] - values[1:(n-1), 1]) /  values[1:(n-1), 1])
iwm.returns$date = as.Date(rownames(values)[2:n])

colnames(iwm.returns) = c('returns','date')

ggplot(iwm.returns, aes(x=date, y=returns,group = 1)) + geom_line() + scale_x_date(breaks = date_breaks("years"),
  labels = date_format('%y')) + xlab('Year')
```

```{r}
k=10
AICs = numeric(length(2:k))

for (i in 2:k) {
  hmm <- depmix(returns ~ 1, family = gaussian(), nstates = i, data=data.frame(iwm.returns))
  hmmfit <- fit(hmm, verbose = FALSE)
  AICs[i-1] = AIC(hmmfit)
}

```

```{r}
bestk = which.min(AICs) +1
bestk
```

```{r}
hmm <- depmix(returns ~ 1, family = gaussian(), nstates = bestk, data=data.frame(iwm.returns))
hmmfit <- fit(hmm, verbose = FALSE)
```

```{r}
summary(hmmfit)
```

```{r}
dat = posterior(hmmfit)
dat$date = iwm.returns$date
dat$ret = iwm.returns$returns
```

```{r}
ggplot(dat, aes(x=date, y=ret,color=factor(state),group = 1)) + geom_line() + scale_x_date(breaks = date_breaks("years"),
  labels = date_format('%y')) + xlab('Year') + scale_color_brewer(palette="Set1")
```

Like with the S&P index, the most volatil bear state appears during the crisis.

\pagebreak
\bibliographystyle{abbrv}
\bibliography{biblio}
